---
title: |
    | ECON 7201
    | Econometrics for Economics and Finance
subtitle: "Lab 3: Quarto"
format: pdf
bibliography: references.bib

execute:
  echo: true
  warning: false
---
\vspace{-1in}

# Preliminaries

1.  Create an R project in RStudio: **File\>New Project...\>New Directory**. Name the directory **econ_7201_lab_3** and save it to your desired location.

-   Note: if you are using your personal computer, you should have a course directory saved in your **Documents** folder with the following sub-directories: **assignments**, **labs**, **notes**. Save this R project under the **labs** folder, i.e., the path tho this lab should be **Documents\>econ_7201\>labs\>econ_7201_lab_3**

2.  Download the lab materials to this directory

# Introduction

Quarto is the publishing software that we will be using for the course. the purpose of today's lecture is to help familiarize yourself with Quarto and its functionalities

# Lab

## Creating a Quarto File

1.  To open a Quarto file, click the **File\>New File\>Quarto Document...**
2.  Name the file "Econ 7201 Lab 3" and choose **PDF**\
3.  The Quarto file will now appear in the source editor (top left pane in RStudio)
4.  Save the file to your lab directory
5.  Delete all default text from “## Quarto” in the file.

# Quarto Basics

## Section Headings

``` markdown
# Section Headings
## Subsection Heading
### Sub-Subsection Heading
```

Render to PDF so students can see the output.

## Formatted Text

-   *italics*: `*italics*`\
-   **bold**: `**bold**`\
-   `code`: `` `code` ``\
-   `[links](lclair.ca)`: [links](https://lclair.ca)

## Lists

**Unordered list:**

``` markdown
- Item 1
- Item 2
  + Item 2a
  + Item 2b
```

**Ordered list:**

``` markdown
1. Item 1
2. Item 2
3. Item 3
   + Item 3a
   + Item 3b
```

# LaTeX Equations

-   Inline math: `$y=mx+b$` gives $y = mx + b$
-   Display math: `$$y=mx+b$$` gives

$$y = mx + b$$

**Examples:**\
- Fraction: `$\frac{a}{b}$` gives $\frac{a}{b}$ - Subscript: `$x_1$` gives $x_1$ - Superscript: `$x^2$` gives $x^2$ - Summation: `$$\sum_{i=1}^N$$` gives $\sum_{i=1}^N$ - Greek letters: `$\alpha$, $\beta$, $\gamma$` gives $\alpha$, \beta\$, $\gamma$

# R Code in Quarto

Three ways to run R code:\
1. Console\
2. R Script file\
3. Code chunk in Quarto

Example R chunks:

```{r}
# Comments with "#"
1 + 1
```

```{r}
# Create a vector
y <- c(1,2,3,4)
y
```

```{r}
# Compute mean of a sequence
z <- seq(from = 1, to = 5, by = 1)
mean(z)
```

# Bibliography

One of Quarto's greatest strength's is its reference management system. 

## .bib Files
To add a bibliography, you need to create a `.bib` file with your references. Each entry in a .bib file begins with an entry type (like `@article`, `@book`, etc.), followed by a citation key and a set of fields in key–value pairs.

| Entry Type | Required Fields | Common Optional Fields |
|----------------------|-----------------------------|---------------------------|
| `@article` | `author`, `title`, `journal`, `year` | `volume`, `number`, `pages`, `month`, `doi`, `url`, `note` |
| `@book` | `author` **or** `editor`, `title`, `publisher`, `year` | `volume`, `series`, `address`, `edition`, `month`, `isbn`, `note` |
| `@incollection` | `author`, `title`, `booktitle`, `publisher`, `year` | `editor`, `pages`, `series`, `address`, `edition`, `month`, `note` |
| `@inproceedings` | `author`, `title`, `booktitle`, `year` | `editor`, `pages`, `organization`, `publisher`, `series`, `address`, `month`, `note` |
| `@techreport` | `author`, `title`, `institution`, `year` | `number`, `type`, `address`, `month`, `note`, `url` |
| `@phdthesis` | `author`, `title`, `school`, `year` | `type`, `address`, `month`, `note`, `url` |
| `@mastersthesis` | `author`, `title`, `school`, `year` | `type`, `address`, `month`, `note`, `url` |
| `@misc` | *none required* (flexible) | `author`, `title`, `howpublished`, `year`, `month`, `note`, `url` |

A sample entry is

```bibtex
@article{smith2020example,
  author  = {Smith, John and Doe, Jane},
  title   = {An Example of a Journal Article},
  journal = {Journal of Examples},
  year    = {2020},
  volume  = {15},
  number  = {3},
  pages   = {45--67},
  doi     = {10.1234/example.doi},
  url     = {https://doi.org/10.1234/example.doi}
}
```

## Citations

To cite a reference from your .bib file use the `[@cite-key]` syntax. For in-text citations, use `@cite-key` syntax. 

## Adding a Bibliography and Citation Style

To add a bibliography, use the `bibliography: ` option in your YAML heading. To add a citation style, download the appropriate `.csl` file to your working directory and add the `csl:` option to your YAML header, e.g.

```quarto
---
title: "Demo"
author: John Doe
bibliography:references.bib
csl: nature.csl
---
```

To add the bibliography to your document, add a section `# References` at the end of your document. Quarto will automatically add the references. 


# Questions

1.  Type the following equations in Quarto/LaTeX:

    a.  $\frac{x^2 + 1}{x}$
    b.  $\sum_{i=1}^n y_i p_i$
    c.  $y = \alpha + \beta x$
    d.  $\frac{\sum_{i=1}^n (y - \bar y)(x - \bar x)}{\sum_{i=1}^n (x - \bar x)^2}$
    e.  $\beta = (X'X)^{-1}X'Y$

2.  Estimate the mean, median, and standard deviation of\
    $Z = (1,2,3,4,5,6,7,8,9,10)$ using `mean()`, `median()`, and `sd()` in R.
    
3. Below is a quote from Smart and Doremus (2023). The references and cite keys for the citations can be found in the refs.bib file in the lab zip folder. 
    a. Change the references in the text below using Quarto syntax.
    b. Add the bibliography to your quarto file.
    c. Add the `apa.csl` to your Quarto YAML heading.
    d. Add the section `# References` at the end of your file
    

> "A final difference between youth and adult use concerns behavioral responses to market features. When consuming addictive goods, youths are more price elastic than adults [@Grossman1997], and evidence from cigarette markets supports that youth are largely reliant on social markets for accessing age-restricted goods (Katzman et al., 2007). Given that legal users’ costs in the formal market affect supply in social markets (Hansen et al., 2013), medical marijuana market size may have an effect on youth use separate from the effect of legalization alone. Consistent with this, when Salomonsen-Sautel et al. (2012) surveyed adolescents engaged in substance use disorder treatment in Denver, Colorado, they found 74% reported using diverted medical marijuana. In 2013, a national survey found that, among high-school seniors, almost 18% of past-year users report access through another person’s medical marijuana “prescription” (@Boyd2015); among high-school seniors in states with MMLs, the share was 34% (Wadley and Barnes, 2013). This suggests substantial diversion from legal to illegal markets that supply adolescents and young adults."
