---
title: |
  | ECON 7201
  | Applied Econometrics
subtitle: "Assignment 3"
format: 
  pdf:
    include-in-header: 
      text: |
        \usepackage{fancyhdr}
        \fancypagestyle{style2}{
        \fancyhf{}
        \fancyhead[R]{Assignment 1}
        \fancyhead[L]{ECON 3201}
        \fancyfoot[C]{\thepage}
        \renewcommand{\headrulewidth}{1pt}
        }
        \pagestyle{style2}
execute: 
  eval: false
  echo: true
---

```{r}
#| label: setup 
#| include: FALSE

#install.packages("tikzDevice")
# Load the tikzDevice library
library(tikzDevice)

# Set the default graphics device to tikz for the entire document
knitr::opts_chunk$set(dev = 'tikz', fig.width = 6, fig.height = 4)

set.seed(42)
```

\vspace{-1in}

## Due Date

**Friday November 7, 2025** at 11:59 PM

## Directions  
Answer each question clearly and concisely. Write equations using LaTeX where appropriate and explain all assumptions in your own words. Submit both a PDF and Quarto file to the nexus assignment portal.
 

# Git and GitHub

1. 
    (a) Create a new R project in your **econ_3201** directory called **assignment_3**.
    (b) Download the assignment PDF and Quarto file the **assignment_3** folder.
    (c) Commit and push the changes to your **econ_3201** repository on [GitHub.com](GitHub.com).


# Conceptual Foundations

1. (**5 pts**) Define the *potential outcomes framework*. Clearly explain what is meant by $Y_i(1)$, $Y_i(0)$, and $D_i$.

2. (**5 pts**) State and explain the **Stable Unit Treatment Value Assumption (SUTVA)**. Why is this assumption critical for causal inference?

3. (**5 pts**) What is the **Conditional Independence Assumption (CIA)**?  Write it formally and explain its meaning in words.  

4. (**5 pts**) Describe the **Overlap (Common Support)** condition and its practical importance.

5. (**5 pts**) Explain why the **Average Treatment Effect on the Treated (ATT)** may differ from the **Average Treatment Effect (ATE)**.


# Identifying Causal Effects (25 points)

1. (**5 pts**) Suppose treatment $ D_i $ is assigned completely at random. Show that  
$$
E[Y_i(1)|D_i=1] - E[Y_i(0)|D_i=0] = ATE
$$
Explain why randomization justifies this equality.

2. (**10 pts**) Suppose treatment depends on observed covariates $ X_i $, such that  
$$
(Y_i(1), Y_i(0)) \perp D_i \mid X_i
$$
Derive an expression for $ATE$ using the **law of iterated expectations**, and interpret it.

3. (**10 pts**) Explain how *selection bias* arises if the CIA fails. Use the decomposition:
$$
E[Y_i|D_i=1] - E[Y_i|D_i=0] = ATE + \text{Selection Bias}.
$$

# Applied Exercise in R (30 points)

Use R to simulate data, estimate treatment effects, and interpret your results. Include all R code and output.


```{r}
#| echo: true
set.seed(123)

n <- 1000
X <- rnorm(n)
D <- ifelse(X + rnorm(n) > 0, 1, 0)
Y0 <- 2 + 0.5 * X + rnorm(n)
Y1 <- Y0 + 3 + 0.5 * X
Y <- ifelse(D == 1, Y1, Y0)
data <- data.frame(Y, D, X)

```


1. (**5 pts**) Compute the difference in average outcomes between treated and untreated. Interpret the result in terms of potential bias. 

2. (**5 pts**)
    (a) Estimate the treatment effect controlling for (X):
$$
Y_i = \alpha + \tau D_i + \beta X_i + \varepsilon_i
$$
    (b) Compare the estimate of (\tau) to the true treatment effect of 3.

3. (**5 pts**) Propensity Score Matching 
    (a) Estimate treatment effects using matching. Search `?matchIt` in the console 
    (b) Compare the matched estimate to OLS and the naïve difference.


4. (**5 pts**) Simulate a violation of the Conditional Independence Assumption by introducing an unobserved confounder ( U_i ):

```{r}
U <- rnorm(n)
D <- ifelse(X + U + rnorm(n) > 0, 1, 0)
Y0 <- 2 + 0.5 * X + 0.5 * U + rnorm(n)
Y1 <- Y0 + 3 + 0.5 * X
Y <- ifelse(D == 1, Y1, Y0)
bias_model <- lm(Y ~ D + X, data = data.frame(Y, D, X))
summary(bias_model)
```

Visualize and discuss how this unobserved confounder biases the estimated treatment effect.

```{r}
plot(X, Y, col = ifelse(D==1, "red", "blue"), pch = 16,
     main = "Outcome vs Covariate with Confounding",
     xlab = "X", ylab = "Y")
legend("topleft", legend = c("Treated", "Untreated"),
       col = c("red", "blue"), pch = 16)
```
